---
title: "Case Study 2 Report 1"
author: "Altamash Rafiq"
date: "29/09/2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data Prep

```{r}
data <- read.table("kellydat.txt", header = TRUE)
library(survminer)
library(survival)
library(plyr)
```

Survival Curves

We made survival curves to illustrate our data using Kaplan-Meier estimation. We used the ggsurvplot command in the survminer R package to make these curves.

The plots are the be interpretted as follows: the y-axis indicates the probability that a given patient has not been scanned yet. This is plotted against time (nctdel) on the x-axis such that at any time x (on the x-axis), the probability that someone has not been scanned yet is y (on the y-axis).

We first plotted a survival curve for the whole data without subsetting it :-

```{r}
fitM <- survfit(Surv(nctdel, fail) ~ 1, data = data)
ggsurvplot(fitM, data = data, risk.table = FALSE) + labs(y = "Probability Unscanned") +
  ggtitle("Survival Curve for Combined Data")
```

As this plot and our calculation suggest that in our sample, the probability of getting scanned in under three hours is lesser than 0.88385.

Next we plotted survival curves for various combinations of symptoms :-

```{r}
fit <- survfit(Surv(nctdel, fail) ~ sn1 + sn2 + sn3 + all4, data = data)
ggsurvplot(fit, data = data, risk.table = FALSE, legend.labs=c("No Symptoms", "All Symptoms", "Only Symptom 3","Only Symptom 2", "Only Symptom 1")) + labs(y = "Probability Unscanned") +
  ggtitle("Survival Curve for Symptoms")
```

This plot and our calculation suggest that in our sample :-

1) If a patient has none of the three symptoms, their probability of getting scanned in under three hours is 0.8237.
2) If a patient has all of the three symptoms, their probability of getting scanned in under three hours is 1.
3) If a patient has only symptom 3, their probability of getting scanned in under three hours is between 0.9499 and 1.
4) If a patient has only symptom 2, their probability of getting scanned in under three hours is between 0.9119 and 0.9296.
5) If a patient has only symptom 1, their probability of getting scanned in under three hours is 0.88517.

As is to be expected, if a patient demonstrates none of the three symptoms, their probability of getting scanned before three hours is the least. Similarly if a patient demonstrates all of the three symptoms, their probability of getting scanned before three hours is 100%. There seem to, however, be a higherarchy of diagnosis between the three symptoms in our sample. The probability of getting scanned before three hours is higher for those with only symptom 3 than those with only symptom 2 which in turn is higher than those with only symptom 1. However, for the majority of the times, the survival curves of the three symptoms appear essentially superimposed.

Next we plotted survival curves for the three races under consideration :-

```{r}
fit1 <- survfit(Surv(nctdel, fail) ~ black + hisp, data = data)
ggsurvplot(fit1, data = data, risk.table = FALSE, legend.labs=c("White", "Hispanic", "Black")) + labs(y = "Probability Unscanned") +
  ggtitle("Survival Curve for Races")
```

This plot and our calculation suggest that in our sample :-

1) If a patient is White, their probability of getting scanned in under three hours is 0.89389.
2) If a patient is Black, their probability of getting scanned in under three hours is between 0.8717 and 0.8859.
3) If a patient is Hispanic, their probability of getting scanned in under three hours is between 0.75 and 0.875. As there are only six hispanics in the sample, this estimate is not reliable and we will not focus on the survival curve for hispanics in our analysis.

As is visible from the graph, at a given time, the probability of a white person getting scanned before a black person is conspicuously higher. The probability of getting scanned before the three hour mark for white people in the sample is also higher than that of black people.

Finally we plotted survival curves for males and females :-

```{r}
fit2 <- survfit(Surv(nctdel, fail) ~ male, data = data)
ggsurvplot(fit2, data = data, risk.table = FALSE, legend.labs=c("Female", "Male")) + labs(y = "Probability Unscanned") +
  ggtitle("Survival Curve for Genders")
```

This plot and our calculation suggest that in our sample :-

1) If a patient is Male, their probability of getting scanned in under three hours is 0.88518 and 0.89284.
2) If a patient is Female, their probability of getting scanned in under three hours is 0.88292.

As is visible from the graph, at a given time, the probability of a female getting scanned before a male is conspicuously higher. However, the probability of males getting scanned before 3 hours is minutely higher than for females.
